---
title: "Modelling the mites"
description: |
  Working out the mite data example.
execute:
  freeze: true
  eval: false
format:
  html:
    code-tools: true
editor_options: 
  chunk_output_type: console
draft: true
---

```{r}
data(mite, package = "vegan")
data("mite.env", package = "vegan")
library(tidyverse)
library(cmdstanr)

# combine data and environment
mite_data_long <- bind_cols(mite.env, mite) |> 
  tibble::rownames_to_column(var= "site_id") |> 
  mutate(site_id = parse_number(site_id)) |> 
  pivot_longer(Brachy:Trimalc2, names_to = "spp", values_to = "abd")


mite_data_long_transformed <- mite_data_long |> 
  mutate(presabs = as.numeric(abd>0),
         # center predictors
         water = (WatrCont - mean(WatrCont)) / 100
         )

# pick a species that has about 50/50 chance 

mite_data_long_transformed |>
  group_by(spp) |>
  summarize(freq = mean(presabs)) |>
  filter(freq > .4 & freq < .6)

```

```{r}

spp_nvec <- unique(mite_data_long_transformed$spp) |> 
  {\(x) set_names(seq_along(x), x)}()

mite_data_long_transformed <- mite_data_long |> 
  mutate(presabs = as.numeric(abd>0),
         # center predictors
         water = (WatrCont - mean(WatrCont)) / 100,
         spp_id = spp_nvec[mite_data_long_transformed$spp]
         )

mite_data_long_transformed |> 
  ggplot(aes(x = water, y = presabs)) + 
  geom_point() + 
  stat_smooth(method = "glm", method.args = list(family = "binomial")) + 
  facet_wrap(~spp)
```

```{r}
mite_bin <- mite
mite_bin[mite_bin>0] <- 1


make_data_list <- function(df){
  mite_pa_list <- 
    with(df, 
         list(
           Nsites = max(site_id),
           Nspp = max(spp_id),
           site_id = site_id,
           spp_id = spp_id,
           x = water,
           y = as.numeric(abd>0)
         )
    )
}
```


## Modelling pres abs

```{r}
#| class-output: stan
presabs_all_spp <- cmdstan_model(
  stan_file = "topics/05_modelling_mites/presabs_all_spp.stan", 
  pedantic = TRUE)

presabs_all_spp
```


```{r}

presabs_all_spp_posterior <- 
  presabs_all_spp$sample(
    data = mite_pa_list, 
    refresh = 1000, parallel_chains = 4
    )
```

### Sampling posterior predictive with `gq`

```{r}

```



## Modelling counts


```{r}
#| class-output: stan
count_all_spp <- cmdstan_model(
  stan_file = "topics/05_modelling_mites/count_all_spp.stan", 
  pedantic = TRUE)

count_all_spp
```


Convert to LONG format all the outputs

```{r}
mite_abd_list <- make_data_list(mite_data_long_transformed)

count_all_spp_post <- count_all_spp$sample(
    data = mite_abd_list, 
    refresh = 200, parallel_chains = 4
    )
```


### PPC using gq

Make predictions with the poisson model using a separate gq block

```{r}
#| class-output: stan
count_all_spp_gq <- cmdstan_model(
  stan_file = "topics/05_modelling_mites/count_all_spp_gq.stan", 
  pedantic = TRUE)

count_all_spp_gq
```

```{r}
count_all_spp_gq_post <- count_all_spp_gq$generate_quantities(count_all_spp_post, data = mite_abd_list)
```


```{r}
yrep_mat <- count_all_spp_gq_post$draws(variables = "yrep") |> 
  posterior::as_draws_matrix()

bayesplot::ppc_dens_overlay(mite_data_long$abd, yrep_mat |> head(50)) + 
  coord_cartesian(xlim = c(0, 20))
```


Does that mean that the probability of getting 1 or higher (ie not zero) from the poisson distribution, will give us a good line for presence-absence?

```{r}
#| class-output: stan
count_all_spp_line_gq <- cmdstan_model(
  stan_file = "topics/05_modelling_mites/count_all_spp_line_gq.stan", 
  pedantic = TRUE)

count_all_spp_line_gq
```

make a new data list with expand.grid

```{r}
new_list <- tibble(water = modelr::seq_range(mite_data_long_transformed$water, n = 10), 
         site_id = 1:10) |> 
  expand_grid(spp_id = unique(mite_data_long_transformed$spp_id), abd = 0) |> 
  make_data_list()


count_all_spp_line_gq$generate_quantities(count_all_spp_post, data = new_list)
```



In this tutorial I want to give one example of a model-buildig process with the mite data. 

* data plots, causal model. Does the causal model suggest the need to control for spacE? there might be collider bias or something, it would be interesting to make that point
* simulated process
* model incidence, straight to hierarchical
* model abundance. note thenumber of zeros. in a ppc check
* calculate the proportion of zeros and plot its densiy, and show the real number of zeros
* 0-inflated, and the similarity to modelling incidence
* correlated effect of the two as a function of some latent trait. 

* do we have traits of mites? 

* adding in effects of environmental variables. 

* ARD? 