---
title: "Summarizing many univariate models"
description: |
  A secret weapon for when you're building hierarchical models.
execute:
  freeze: false
format:
  html:
    code-tools: true
editor_options:
  chunk_output_type: console
---

We've already looked at univariate models. When we fit the same model to multiple different groups, we don't expect the same values for all the coefficients. Each thing we are studying will respond to the same variable in different ways.

Hierarchical models represent a way to *model* this variation, in ways that range from simple to complex.

Before we dive in with hierarchical structure, let's build a bridge between these two approaches.

This is useful to help us understand what a hierarchical model does.

However it is also useful from a strict model-building perspective -- so useful that [Andrew Gelman calls it a "Secret Weapon"](https://statmodeling.stat.columbia.edu/2005/03/07/the_secret_weap/)

## Loading models and data

```{r setup}
library(rstan)
rstan_options("auto_write" = TRUE)
options(mc.cores = parallel::detectCores())

data(mite, package = "vegan")
data("mite.env", package = "vegan")
data("mite.xy", package = "vegan")
```

And some quick data restructuring to combine both.

```{r datasetup}
# Combine data and environment
spp_cols_ce <- colnames(mite)
mite_combined_ce <- cbind(mite.env, mite)

mite_data_long <- do.call(rbind, lapply(spp_cols_ce, function(sp) {
  data.frame(WatrCont = mite_combined_ce$WatrCont,
             spp = sp,
             abd = mite_combined_ce[[sp]])
}))
rownames(mite_data_long) <- NULL
```

To keep things simple and univariate, let's consider only water concentration as an independent variable.

First, a quick word about centring and scaling a predictor variable. In short, think of the biology of your system before transforming the data. Furthermore, make sure you understand how these transformation may influence the model parameters. 

Here, I decide to centre the predictor by subtracting the mean. This changes the *intercept* of my linear predictor. It becomes the mean log-odds of occurrence when the water content is average.

I decided not to perform additional transformation on the water content because I still want to retain the units associated to this explanatory variable. This may change if I want to build a model that include multiple explanatory variables with different units. In this case, we may want to apply a transformation on the explanatory variables that removes the effect of the units. 

```{r smoothglm}
mite_data_long_transformed <- mite_data_long
mite_data_long_transformed$presabs <- as.numeric(mite_data_long_transformed$abd > 0)
mite_data_long_transformed$water <- (mite_data_long_transformed$WatrCont -
                                     mean(mite_data_long_transformed$WatrCont))

u_spp_ce <- unique(mite_data_long_transformed$spp)

nc_ce <- 5
nr_ce <- ceiling(length(u_spp_ce) / nc_ce)

par(mfrow = c(nr_ce, nc_ce), 
    mar = c(2, 2, 1.5, 0.3))

for (sp in u_spp_ce) {
  df_sp <- mite_data_long_transformed[mite_data_long_transformed$spp == sp, ]
  
  plot(df_sp$water, 
       df_sp$presabs,
       pch = 19, 
       cex = 0.4, 
       col = adjustcolor("steelblue", 0.4),
       xlab = "",
       ylab = "", 
       main = sp, 
       cex.main = 0.7)
  
  fit_sp <- glm(presabs ~ water, 
                data = df_sp, 
                family = binomial)
  
  x_s    <- seq(min(df_sp$water), 
                max(df_sp$water), 
                length.out = 100)
  
  lines(x_s, predict(fit_sp, 
                     newdata = data.frame(water = x_s), 
                     type = "response"),
        col = "steelblue", 
        lwd = 1.5)
}
```

An important things to notice about this figure is that there is a ton of variation in how different species respond to water!

```{r make_mite_glms}
u_spp_glm <- unique(mite_data_long_transformed$spp)

mite_many_glms_list <- lapply(u_spp_glm, function(s) {
  df <- mite_data_long_transformed[mite_data_long_transformed$spp == s, ]
  glm(presabs ~ water, family = "binomial", data = df)
})

names(mite_many_glms_list) <- u_spp_glm

mite_many_glm_coefs <- do.call(rbind, lapply(u_spp_glm, function(s) {
  cf <- coef(summary(mite_many_glms_list[[s]]))
  data.frame(spp = s,
             term = rownames(cf),
             estimate = cf[, "Estimate"],
             std.error = cf[, "Std. Error"],
             stringsAsFactors = FALSE)
}))
rownames(mite_many_glm_coefs) <- NULL
```

:::{.callout-note}
## Split-Apply-Combine

To explore this kind of thinking, we are going to use an approach sometimes called ["split-apply-combine"](https://vita.had.co.nz/papers/plyr.pdf)

There are many possible ways to do this in practice. Above, we used `lapply` to carry out the split-apply-combine pattern. That is, we loop over species, fit one GLM per species, and collect the coefficients with `coef(summary())`.
:::

```{r coef_plot}
int_df <- mite_many_glm_coefs[mite_many_glm_coefs$term == "(Intercept)", ]
slp_df <- mite_many_glm_coefs[mite_many_glm_coefs$term == "water", ]

par(mfrow = c(1, 2), mar = c(4, 5, 2, 0.5))

for (df_t in list(int_df, slp_df)) {
  xlim_t <- range(c(df_t$estimate - 2*df_t$std.error, 
                    df_t$estimate + 2*df_t$std.error))
  
  plot(df_t$estimate, 
       seq_along(df_t$spp),
       pch = 19, 
       yaxt = "n",
       xlim = xlim_t,
       xlab = "Estimate", 
       ylab = "",
       cex = 0.75,
       main = df_t$term[1])
  
  abline(v = 0, col = "red")
  
  segments(df_t$estimate - df_t$std.error, 
           seq_along(df_t$spp),
           df_t$estimate + df_t$std.error,
           seq_along(df_t$spp))
  
  axis(2, at = seq_along(df_t$spp),
       labels = df_t$spp, 
       las = 1, 
       cex.axis = 0.6)
}
```

As you can see, some of these estimates are high, others low. We could also plot these as histograms to see this distribution.

```{r secretweapon_hist}
par(mfrow = c(1, 2))

hist(int_df$estimate, 
     xlab = "Estimate",
     main = "(Intercept)")

hist(slp_df$estimate, 
     xlab = "Estimate",
     main = "water")
```

Once again, the two parameters of this model represent :

-   *Intercept* The probability (in log-odds) of a species being present at the average water concentration. some species are common, others are rare.
-   *water* this is the change in probability (in log-odds) as water increases by one centilitre per litre of substrate.

## Let's do it in Stan

The above `lapply` approach is useful for quick exploration, but we can also do the same procedure in Stan.

```{r}
#| class-output: stan
all_species_unpooled <- stan_model(
  file = "topics/correlated_effects/all_species_unpooled.stan")

all_species_unpooled
```

Let's fit this model by passing in the data:

```{r}
mite_bin <- mite
mite_bin[mite_bin>0] <- 1

mite_pa_list <- list(Nsites = nrow(mite_bin),
                     S = ncol(mite_bin),
                     x = with(mite.env, 
                              (WatrCont - mean(WatrCont))),
                     y = as.matrix(mite_bin))

all_species_unpooled_posterior <- rstan::sampling(all_species_unpooled,
                                                  data = mite_pa_list,
                                                  refresh = 1000, 
                                                  chains = 4)
```

Now let's try to plot this:

```{r}
#| warning: false
# start by looking at the names of variables

# Helper: draw per-species ribbon plots from a Stan fit
draw_spp_ribbons <- function(fit, water_seq = seq(-300, 400, length.out = 100),
                             spp_names = colnames(mite_bin),
                             fill_col = "green4",
                             add_data = FALSE) {
  
  draws_int <- rstan::extract(fit, "intercept")$intercept
  draws_slp <- rstan::extract(fit, "slope")$slope
  
  n_spp <- ncol(draws_int)
  nc <- 5
  nr <- ceiling(n_spp / nc)
  
  par(mfrow = c(nr, nc), 
      mar = c(2, 2, 1.5, 0.3))
  
  for (j in seq_len(n_spp)) {
    
    pm <- matrix(0, nrow(draws_int), length(water_seq))
    
    for (wi in seq_along(water_seq)){
      pm[, wi] <- plogis(draws_int[, j] + draws_slp[, j] * water_seq[wi])
    }
    
    q_pp <- apply(pm, 
                  2, 
                  quantile, 
                  probs = c(0.025, 0.25, 0.5, 0.75, 0.975))
    
    plot(0,
         0, 
         type = "n", 
         xlim = range(water_seq), 
         ylim = c(-0.05, 1.05),
         xlab = "", 
         ylab = "", 
         main = spp_names[j], 
         cex.main = 0.8)
    
    polygon(c(water_seq, 
              rev(water_seq)), 
            c(q_pp[1,], 
              rev(q_pp[5,])),
            col = adjustcolor(fill_col, 0.15), 
            border = NA)
    
    polygon(c(water_seq,
              rev(water_seq)), 
            c(q_pp[2,], 
              rev(q_pp[4,])),
            col = adjustcolor(fill_col, 0.30),
            border = NA)
    
    lines(water_seq, 
          q_pp[3,], 
          col = fill_col, 
          lwd = 1.5)
    
    if (add_data) {
      sp_df <- mite_data_long_transformed[mite_data_long_transformed$spp == spp_names[j], ]
      
      colo <- col2rgb(fill_col)
      colPts <- rgb(colo[1]/255,
                    colo[2]/255,
                    colo[3]/255,
                    alpha = 0.3)
      
      points(sp_df$water, 
             sp_df$presabs,
             pch = 21, 
             cex = 0.5,
             bg = colPts,
             col = colPts)
    }
  }
}

draw_spp_ribbons(all_species_unpooled_posterior, 
                 fill_col = "green4",
                 add_data = TRUE)
```

:::{.callout-tip}
### EXERCISE
1) Add hierarchy to both the slope **and** the intercept of this model. Note that the `lme4` syntax would be `y ~ 1 + water + (1 | species) + (0 + water | species)`

2) Make a plot of the slope coefficients from both models, with and without hierarchy. Are they the same? How are they different?

3) Make a posterior prediction of species richness in these communities.
:::

:::{.callout-note collapse="true"}
### SOLUTION

First we rewrite the Stan model from above, replacing the standard deviations with parameters. We do this for the priors on the intercepts and slopes _separately_.

```{r}
#| class-output: stan
all_species_partpooled <- stan_model(
  file = "topics/correlated_effects/all_species_partpooled.stan")

all_species_partpooled
```

Sample the model

```{r}
all_species_partpooled_posterior <- sampling(all_species_partpooled,
                                             data = mite_pa_list,
                                             refresh = 1000, # What is reported
                                             chains = 4)
```

Plot posterior predictions of trend lines, just as before:

```{r}
#| fig-height: 9
#| fig-width: 7
draw_spp_ribbons(all_species_partpooled_posterior,
                 fill_col = "red3",
                 add_data = TRUE)
```

#### Comparing slope estimates of both models


```{r}
draws_slp_up <- rstan::extract(all_species_unpooled_posterior, 
                               "slope")$slope
draws_slp_pp <- rstan::extract(all_species_partpooled_posterior,
                               "slope")$slope
spp_names_ce <- colnames(mite_bin)

q_up <- apply(draws_slp_up, 
              2, 
              quantile,
              probs = c(0.025, 0.25, 0.5, 0.75, 0.975))

q_pp <- apply(draws_slp_pp, 
              2, 
              quantile,
              probs = c(0.025, 0.25, 0.5, 0.75, 0.975))

ord <- order(q_pp[3, ])
n_sp <- length(spp_names_ce)

plot(0,
     0, 
     type = "n",
     xlim = c(0.5, n_sp + 0.5), 
     ylim = range(c(q_up, q_pp)),
     xlab = "",
     ylab = "Slope (water)",
     xaxt = "n")

axis(1,
     at = seq_len(n_sp), 
     labels = spp_names_ce[ord], 
     las = 2, 
     cex.axis = 0.6)

abline(h = 0, 
       lty = 2, 
       col = "grey60")

for (j in seq_len(n_sp)) {
  oj <- ord[j]
  for (xoff in list(list(x = j - 0.15, q = q_up[, oj], col = "steelblue"),
                    list(x = j + 0.15, q = q_pp[, oj], col = "orange"))) {
    
    segments(xoff$x, 
             xoff$q[1], 
             xoff$x, 
             xoff$q[5], 
             col = xoff$col, 
             lwd = 1)
    
    segments(xoff$x, 
             xoff$q[2],
             xoff$x, 
             xoff$q[4], 
             col = xoff$col,
             lwd = 3)
    
    points(xoff$x,
           xoff$q[3],
           pch = 19,
           col = xoff$col)
  }
}

legend("topleft",
       legend = c("non-hierarchical", "hierarchical"),
       col = c("steelblue", 
               "orange"), 
       pch = 19)

```

#### Posterior distribution of species richness

We can always calculate things out of the posterior distribution, and get a new distribution which reflects the uncertainty in all our parameter estimates.
Here I'm suggesting we calculate the relationship between **species richness** and water concentration

```{r}
draws_int_pp_S <- rstan::extract(all_species_partpooled_posterior, 
                                 "intercept")$intercept

draws_slp_pp_S <- rstan::extract(all_species_partpooled_posterior, 
                                 "slope")$slope

water_seq_S <- seq(-8, 6, length.out = 10)
idx_S <- sample(nrow(draws_int_pp_S), 
                min(700, 
                    nrow(draws_int_pp_S)))

S_mat <- matrix(0, 
                length(idx_S), 
                length(water_seq_S))

for (ki in seq_along(idx_S)) {
  k <- idx_S[ki]
  for (wi in seq_along(water_seq_S))
    S_mat[ki, wi] <- sum(plogis(draws_int_pp_S[k, ] + draws_slp_pp_S[k, ] *
                                  water_seq_S[wi]))
}

q_S <- apply(S_mat, 
             2, 
             quantile, 
             probs = c(0.025, 0.25, 0.5, 0.75, 0.975))

plot(0,
     0,
     type = "n",
     xlim = range(water_seq_S), 
     ylim = range(q_S),
     xlab = "Water (centered)",
     ylab = "Expected species richness")

polygon(c(water_seq_S, rev(water_seq_S)), 
        c(q_S[1,], rev(q_S[5,])),
        col = adjustcolor("steelblue",
                          0.15),
        border = NA)

polygon(c(water_seq_S, rev(water_seq_S)), 
        c(q_S[2,], rev(q_S[4,])),
        col = adjustcolor("steelblue", 0.30),
        border = NA)

lines(water_seq_S, 
      q_S[3,],
      col = "steelblue", 
      lwd = 2)
```

:::
