---
title: "Extra course content"
description: |
  A collection of topics that didn't fit in the course schedule but are still useful and interesting!
execute:
  freeze: true
comments:
  hypothesis: true
format:
  html:
    code-tools: true
editor_options: 
  chunk_output_type: console
---

## Load packages and data

```{r}
library(palmerpenguins)
library(tidyverse)
# library(cmdstanr)
suppressPackageStartupMessages(library(rstan))
rstan_options("auto_write" = TRUE)
options(mc.cores = parallel::detectCores())
library(tidybayes)
```



## Simulation workout! 

1) make a histogram of 500 numbers from a distribution! 
   * normal 
   * poisson
   * **EXTRA** try a new one, like beta, gamma, lognormal
   
2) make a histogram of poisson observations, using the classic **`log()` link function**. 

$$
\begin{align}
y &\sim \text{Poisson}(e^a) \\
a &\sim \text{Normal}(??, ??)
\end{align}
$$

:::{.callout-note collapse="true"}
### TIP
```{r}
## sample poisson variables like this:
rpois(500, exp(3))
```
:::

3)  make a histogram of Binomial observations, using the **inverse logit link function**

$$
\begin{align}
y &\sim \text{Binomial}\left(\frac{1}{1+e^{-a}}, N \right) \\
a &\sim \text{Normal}(??, ??)
\end{align}
$$

Here's a plot of the link function, to help you think about it:

```{r}
curve(1 / (1 + exp(-x)), xlim = c(-3, 3), ylim = c(0, 1))
```


:::{.callout-note collapse="true"}
### TIP
```{r}
a <- rnorm(1, mean = 0, 1)
hist(rbinom(n = 500, size = 50, prob = 1 / (1 + exp(-a))))
```
:::

