---
title:  Fitting an intercept-only model
description: |
  Where is the variation
execute:
  freeze: true
eval: false
format:
  html:
    code-tools: true
editor_options: 
  chunk_output_type: console
---

## Variance partitioning with hierarchical models

Its very useful to have an idea of where the variation is in a dataset, as a guide to model building and future data collection. Before collecting information on independent variables that you think might explain variation -- first find out how much variation there is! 

Random intercepts give us a very handy way to investigate this: the so-called "intercept only" model. 
At this point in the course we now have all the tools needed to build it. There are two steps: first we build a model with no predictors at all, only a random intercept for every grouping variable in the data (e.g. species, sites, years, regions). Then we examine the relative magnitudes of the standard deviations to see which is relatively larger or smaller.

## Abundance of mites in different samples

We're going to build this by extending the observation level random effect model from the previous section. Remember that observation-level random effects are mostly useful for Poisson distributions! If you want to extend this model to other kinds of data, remember to remove that part of it.

### Mathematical model

$$
\begin{align}
\text{Abundance}_i &\sim \text{Poisson}(\lambda_i) \\
\log{\lambda_i} &\sim \mu + \beta_{\text{sample}[i]} + \beta_{\text{species[i]}} + \beta_i\\
\mu &\sim \text{Normal}(3, 1)\\
\beta_{\text{sample}} &\sim \text{Normal}(0,  \sigma_{\text{samp}})\\
\beta_{\text{species}} &\sim \text{Normal}(0, \sigma_{\text{species}})\\
\beta_i &\sim \text{Normal}(0,                \sigma_{\text{obs}}) \\
\sigma_{\text{samp}}    &\sim \text{Exponential}(3)\\
\sigma_{\text{species}} &\sim \text{Exponential}(3)\\
\sigma_{\text{obs}}     &\sim \text{Exponential}(3)
\end{align}
$$

### Load packages and prepare data

```{r}
suppressPackageStartupMessages(library(dplyr))
library(ggplot2)
library(tidyr)
# library(cmdstanr)
suppressPackageStartupMessages(library(rstan))
rstan_options("auto_write" = TRUE)
options(mc.cores = parallel::detectCores())
library(tidybayes)


data("mite", package = "vegan")
```

```{r}
spp_names <- colnames(mite)
spp_names <- setNames(1:ncol(mite), colnames(mite))


mite_long <- mite |> 
  mutate(site_id = seq_len(nrow(mite))) |> 
  tidyr::pivot_longer(-site_id,
                      names_to = "spp",
                      values_to = "abd") |> 
  dplyr::mutate(spp_id = spp_names[spp])

knitr::kable(head(mite_long))
```

```{r}
#| class-output: stan

spp_site_obs_intercepts <- stan_model(
  file = "topics/intercept_only/spp_site_obs_intercepts.stan",
  model_name = "spp_site_obs_intercepts")

spp_site_obs_intercepts
```

Now we can sample this model. 

:::{.callout-warning}
## Warning: irresponsible statistics
I'm sampling only 2 chains below, for illustration purposes only! use more chains in your research.
:::

```{r eval=TRUE}
spp_site_obs_intercepts_samp <- sampling(
  spp_site_obs_intercepts,
  data = list(
    N = nrow(mite_long),
    N_spp = max(mite_long$spp_id),
    spp_id = mite_long$spp_id,
    N_sites = max(mite_long$site_id),
    site_id = mite_long$site_id,
    abd = mite_long$abd
  ),chains = 2)

# spp_site_obs_intercepts_samp$save_object("topics/intercept_only/spp_site_obs_intercepts.rds")
```

## Exploring the model output

```{r}
rstan::check_divergences(spp_site_obs_intercepts_samp)
```

The sampler was just fine! Note that we have just estimated many more parameters than observations.

## Plot the standard deviations

This just takes one line of bayesplot code.

```{r}
bayesplot::mcmc_areas(spp_site_obs_intercepts_samp,
                      regex_pars ='sigma', 
                      border_size = 0.5)
```

We can also do this as a data frame:

```{r}
sigma_post <- rstan::extract(spp_site_obs_intercepts_samp,
  pars = c("sigma_spp", "sigma_sites", "sigma_obs"))

sigma_post_df <- posterior::as_draws_df(sigma_post)

sigma_post_df |> 
  pivot_longer(starts_with("sigma"), 
               names_to = "sigma", 
               values_to = "value") |> 
  ggplot(aes(x = value)) + 
  geom_histogram() + 
  facet_wrap(~sigma)
```


### Calculate the posterior distribution of average abundance for each species

To do this we need to extract the average $mu$ and add it to the species effects, $\beta_{\text{species}}$

```{r eval=FALSE, include=FALSE}
spp_avg_effects_df <- rstan::extract(spp_site_obs_intercepts_samp, 
  pars = c("mu", "spp_effects")) |> 
  posterior::as_draws_df()

spp_avg_effects_df |> 
  select(mu, starts_with("spp_effects")) |> 
  mutate(row_id = seq_along(mu)) |> 
  pivot_longer(-c("mu", "row_id"), 
               names_to = "parname", 
               values_to = "spp_effect") |> 
  mutate(spp_avg = mu + spp_effect) |> 
  ggplot(aes(x = spp_avg, group = parname)) + 
  geom_density()
```


```{r}
spp_mu_rvars <- spp_site_obs_intercepts_samp |> 
  tidybayes::spread_rvars(mu, spp_effects[spp_id])
```

```{r}
spp_mu_rvars |> 
  mutate(spp_avg = mu + spp_effects) |> 
  ggplot(aes(dist = spp_avg, group = spp_id)) + 
  tidybayes::stat_slab(col = "black") + 
  coord_flip() + 
  theme_minimal()
```

