---
title: "Univariate regression"
description: |
  The shortest route to science is a straight line.
execute:
  freeze: false
comments:
  hypothesis: true
format:
  html:
    code-tools: true
editor_options:
  chunk_output_type: console
---

## Load packages and data

```{r}
library(palmerpenguins)
library(rstan)
rstan_options("auto_write" = TRUE)
options(mc.cores = parallel::detectCores())
```

## Statistical models of Penguin bill morphology.

We'll be studying the relationship between two numbers about penguin bills.
Specifically, we'll ask **"Are longer bills also deeper?"**

This question might not be the most interesting ecologically, but it is a great chance to practice some interesting stats.

Let's begin with plotting the data :

```{r}
#| fig-cap: Bill depth (mm) as predicted by bill length (mm) across the entire `palmerpenguins` dataset.

# Remove NAs
penguins_clean <- na.omit(penguins[, c("bill_length_mm", "bill_depth_mm")])

# Plot data
plot(penguins_clean$bill_length_mm,
     penguins_clean$bill_depth_mm,
     xlab = "Bill length (mm)",
     ylab = "Bill depth (mm)",
     pch = 19, 
     cex = 0.7)

# Regression
reg <- lm(bill_depth_mm ~ bill_length_mm,
          data = penguins_clean)

# Regression lines
abline(reg, 
       col = "steelblue",
       lwd = 3)
```

Let's write a simple statistical model for these data:

$$
\begin{align}
\text{Bill depth}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1\times\text{Bill length}_i \\
\beta_0 &\sim \text{Normal}(??) \\
\beta_1 &\sim \text{Normal}(??) \\
\sigma &\sim \text{Exponential}(??)
\end{align}
$$

What should our priors be? 

Before we can answer that question, we have a more important question:

:::{.callout-warning}
# WHERE IS ZERO?
Specifically, what does it mean *biologically* when bill lengths is at 0 ? Does it make sense? 
:::

If we fit a model like this **without** thinking about the location of zero, we get some pretty silly answers:

```{r}
coef(lm(bill_depth_mm ~ bill_length_mm, data = penguins))
```

When the value of bill length is 0, the average of the response is the intercept:

$$
\begin{align}
\mu_i &= \beta_0 + \beta_1\times\text{Bill length}_i \\
\mu_i &= \beta_0 + \beta_1\times0 \\
\mu_i &= \beta_0 \\
\end{align}
$$

But, if we take the data as we found it, we're going to be talking about $\beta_0$ as the depth of a penguin's bill _when the bill has 0 length!_
Clearly that isn't a very meaningful value. From the point of view of setting priors and interpreting coefficients, it helps a lot to set a meaningful 0.

A very common choice is to **subtract the average** from your independent variable, so that penguins with an average bill length now have a mean of 0:

$$
\begin{align}
\text{Bill depth}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1\times(\text{Bill length}_i  - \overline{\text{Bill length}})\\
\beta_0 &\sim \text{Normal}(??) \\
\beta_1 &\sim \text{Normal}(??)
\end{align}
$$

Now $\beta_0$ means the average _bill depth_ at the average _bill length_.

Using this "new" data, it becomes easier to think about priors:

$$
\begin{align}
\text{Bill depth}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1\times(\text{Bill length}_i  - \overline{\text{Bill length}})\\
\beta_0 &\sim \text{Normal}(17,2) \\
\beta_1 &\sim \text{Normal}(0,.5) \\
\sigma &\sim \text{Exponential}(0.5)
\end{align}
$$

:::{.callout-note}
## Exercise

What continuous predictors have you used in your analysis? How would you find a biologically meaningful zero? Think about how you would centre time, age, mass, fitness etc.
:::

## Prior predictive simulations

Armed with this model, it becomes much easier to think about prior predictions.

We'll make a bunch of lines derived from the equation above. There's two steps:

1. Centre the predictor
2. Make up a vector that goes from the minimum to the maximum of the predictor. This is just for convenience!

```{r}
bill_len_centered <- with(penguins_clean,
                          bill_length_mm - mean(bill_length_mm))

## make up a short vector
some_bill_lengths <- seq(from = min(bill_len_centered),
                         to = max(bill_len_centered),
                         length.out = 10)
```

:::{.callout-warning}
## Shortcuts to these common tasks

These tasks are so common that they are automated in helper functions.

For centering predictors, see the `?scale` function `?scale`

For creating a short vector over the range of a predictor, use `seq(min(x), max(x), length.out = n)`.
:::

To simulate, we'll use some matrix algebra, as we saw in lecture:

```{r}
# Parameters to simulate
slopes <- rnorm(7, 0, .5)
inters <- rnorm(7, 17, 2)

# Basis of the regression
X <- cbind(1, some_bill_lengths)
B <- rbind(inters, slopes)

# Bill length
knitr::kable(head(X))

# Regression parameters
knitr::kable(head(B))

# Simulation for mu 
prior_mus <- X %*% B

# Draw results
matplot(x = some_bill_lengths,
        y = prior_mus, type = "l")
```

:::{.callout-note}
## Exercise

Copy the code above. Increase the number of simulations and study the structure of the priors. Which prior are too wide? Which are too narrow?
:::

### Simulating Observations

There are always at least TWO kinds of predictions we can be thinking about:

1. Predicted averages. This is often called a "confidence" interval for a regression line.
2. Predicted observations. This is often called a "prediction" interval.

We can use the full model to simulate observations!

```{r}
# Simulate parameters
slopes <- rnorm(7, 0, .5)
inters <- rnorm(7, 17, 2)
sigmas <- rexp(7, rate = 0.3)

# Basis of the regression
X <- cbind(1, some_bill_lengths)
B <- rbind(inters, slopes)

# Simulation for mu 
prior_mus <- X %*% B

# Result object for the simulated mu 
prior_obs <- matrix(0, nrow = nrow(prior_mus), ncol = ncol(prior_mus))

# Simulate observation
for (j in 1:ncol(prior_obs)) {
  prior_obs[,j] <- rnorm(n = nrow(prior_mus),
                         mean = prior_mus[,j],
                         sd = sigmas[j])
}

# Plot results
matplot(x = some_bill_lengths,
        y = prior_obs, type = "p")
```

Base R version with small multiples:

```{r}
set.seed(42)
n_sim <- 7

# Simulate parameters
slopes <- rnorm(n_sim, 0, .5)
inters <- rnorm(n_sim, 17, 2)
sigmas <- rexp(n_sim, rate = 0.2)

# Generate X
x_sim  <- seq(-10, 10, length.out = 6)

# Plot structure
par(mfrow = c(2, 4), mar = c(3, 3, 2, 1))

for (i in seq_len(n_sim)) {
  # Build regression
  avg <- inters[i] + slopes[i] * x_sim
  
  # Sample observations from regression average
  obs <- rnorm(length(avg), mean = avg, sd = sigmas[i])
  
  # Plot regression result
  plot(x_sim,
       avg, 
       type = "l",
       col = i + 1,
       lwd = 2,
       ylim = range(c(avg, obs)), 
       xlab = "x",
       ylab = "value",
       main = paste("Sim", i))
  # Plot points
  points(x_sim, obs, pch = 21, bg = i + 1, col = i + 1)
}
```

:::{.callout-tip}
### EXERCISE
Pick one of the two simulations above and modify it. Here are some suggested modifications:

* Experiment with priors that are "too narrow" or "too wide".
* Try a different distribution than the one used
* Instead of bill size, imagine that we are applying this model to YOUR data. What would you change?
:::

## Linear regression in Stan

Now let's write this model with Stan.
We'll begin with a simple model that has no posterior predictions:

```{r}
#| class-output: stan
normal_regression_no_prediction <- stan_model(
  file = "topics/02_regression/normal_regression_no_prediction.stan",
  model_name = "normal_regression_no_prediction")

normal_regression_no_prediction
```

In order to get the posterior, we need to put our data in Stan. We follow the same steps as previously:

* Remember to remove NAs first!
* arrange the data in a list
* pass the data to a Stan model to estimate.

```{r}
penguins_no_NA <- na.omit(penguins[, c("bill_length_mm", "bill_depth_mm", "species")])

penguins_no_NA$bill_length_center <- penguins_no_NA$bill_length_mm -
                                     mean(penguins_no_NA$bill_length_mm)

## Assemble data list
data_list <- with(penguins_no_NA,
                  list(N = length(bill_length_center),
                       bill_len = bill_length_center,
                       bill_dep = bill_depth_mm))

## Run the sampler, using the compiled model
normal_reg_no_pred <- sampling(
  object = normal_regression_no_prediction,
  data = data_list,
  refresh = 0)

summary(normal_reg_no_pred)$summary
```

```{r}
bayesplot::mcmc_areas(normal_reg_no_pred,
                      pars = c("slope", 
                               "intercept", 
                               "sigma"))
```

:::{.callout-tip}
### EXERCISE
**Discussion** : Look only at the posterior distribution of the slope above.
Do we have evidence that there's a relationship between bill length and bill depth.
:::


## Posterior predictions in R

We can calculate a posterior prediction line directly in R for these data.
I'll show each step in this workflow separately:

```{r}
draws_inspect <- rstan::extract(normal_reg_no_pred, 
                                pars = c("slope", "intercept", "sigma"))
str(draws_inspect)
```

`rstan::extract()` pulls the posterior draws for each parameter into a named list — e.g. `draws_inspect$slope`, `draws_inspect$intercept` are numeric vectors with one value per MCMC draw.

Next we combine these posteriors with a vector of observations to make a posterior distribution of LINES:


```{r}
draws_reg <- rstan::extract(normal_reg_no_pred, pars = c("slope", "intercept"))
x_seq     <- seq(-15, 15, length.out = 5)

# mu_draws: matrix (rows = MCMC draws, cols = x-values)
mu_draws  <- outer(as.vector(draws_reg$slope), x_seq) + as.vector(draws_reg$intercept)
str(mu_draws)
```

Finally, let's plot these:

```{r}
q <- apply(mu_draws, 2, quantile, probs = c(0.025, 0.25, 0.5, 0.75, 0.975))

plot(penguins_no_NA$bill_length_center, penguins_no_NA$bill_depth_mm,
     xlab = "Bill length (centered, mm)", ylab = "Bill depth (mm)", pch = 19, cex = 0.5)

# 95% Confidence intervals
polygon(c(x_seq, rev(x_seq)), c(q[1,], rev(q[5,])),
        col = adjustcolor("steelblue", 0.15), border = NA)

# 50% Confidence intervals
polygon(c(x_seq, rev(x_seq)), c(q[2,], rev(q[4,])),
        col = adjustcolor("steelblue", 0.30), border = NA)

# Model
lines(x_seq, q[3,], col = "steelblue", lwd = 2)
```

#### Using posterior draws individually

The above workflow makes a nice figure, but perhaps it helps to see the individual lines to understand what is happening here.

We can get these by sampling rows directly from the extracted draws:

```{r}
set.seed(42)
draw_idx <- sample(length(draws_reg$slope), 12)
draws12 <- data.frame(.draw = draw_idx,
                      slope = draws_reg$slope[draw_idx],
                      intercept = draws_reg$intercept[draw_idx])

knitr::kable(draws12)
```

```{r}
plot(penguins_no_NA$bill_length_center, penguins_no_NA$bill_depth_mm,
     xlab = "Bill length (centered, mm)", ylab = "Bill depth (mm)", pch = 19, cex = 0.5)
for (i in seq_len(nrow(draws12))) {
  lines(x_seq, draws12$intercept[i] + draws12$slope[i] * x_seq,
        col = adjustcolor("steelblue", 0.6))
}
```


## Posterior predictions in Stan

We can also make these posterior predictions in Stan.

First, compile the model:

```{r}
#| class-output: stan
normal_regression <- stan_model(file = here::here("topics/02_regression/normal_regression.stan"))

normal_regression
```

Then prepare the data and feed it into the `sampling()` function.

```{r}
penguins_no_NA <- na.omit(penguins[, c("bill_length_mm", 
                                       "bill_depth_mm", 
                                       "species")])

penguins_no_NA$bill_length_center <- penguins_no_NA$bill_length_mm - 
                                     mean(penguins_no_NA$bill_length_mm)

data_list <- with(penguins_no_NA,
     list(N = length(bill_length_center),
          bill_len = bill_length_center,
          bill_dep = bill_depth_mm,
          npost = 6,
          pred_values = seq(min(penguins_no_NA$bill_length_center),
                            max(penguins_no_NA$bill_length_center), length.out = 6)
          ))

bill_norm_reg <- sampling(normal_regression,
                          data = data_list,
                          refresh = 0)
```


```{r}
#| layout-ncol: 2
draws_avg <- rstan::extract(bill_norm_reg, "post_bill_dep_average")$post_bill_dep_average
draws_obs <- rstan::extract(bill_norm_reg, "post_bill_dep_obs")$post_bill_dep_obs
x_pred    <- data_list$pred_values

q_avg <- apply(draws_avg, 2, quantile, probs = c(0.025, 0.25, 0.5, 0.75, 0.975))
q_obs <- apply(draws_obs, 2, quantile, probs = c(0.025, 0.25, 0.5, 0.75, 0.975))

y_range <- range(penguins_no_NA$bill_depth_mm)

# Average response ()
plot(penguins_no_NA$bill_length_center,
     penguins_no_NA$bill_depth_mm,
     xlab = "Bill length (centered, mm)", 
     ylab = "Bill depth (mm)",
     main = "Average response", 
     pch = 19, 
     cex = 0.5, 
     ylim = y_range)

# 95 % interval
polygon(c(x_pred, rev(x_pred)), 
        c(q_avg[1,], rev(q_avg[5,])),
        col = adjustcolor("darkgreen", 0.15),
        border = NA)

# 50 % interval
polygon(c(x_pred, rev(x_pred)),
        c(q_avg[2,], rev(q_avg[4,])),
        col = adjustcolor("darkgreen", 0.30),
        border = NA)

lines(x_pred, q_avg[3,], 
      col = "darkgreen",
      lwd = 2)

# Predicted observations
plot(penguins_no_NA$bill_length_center,
     penguins_no_NA$bill_depth_mm,
     xlab = "Bill length (centered, mm)",
     ylab = "Bill depth (mm)",
     main = "Predicted observations",
     pch = 19, 
     cex = 0.5, 
     ylim = y_range)

# 95 % Interval
polygon(c(x_pred, rev(x_pred)),
        c(q_obs[1,], rev(q_obs[5,])),
        col = adjustcolor("darkgreen", 0.15),
        border = NA)

# 50 % Interval
polygon(c(x_pred, rev(x_pred)), 
        c(q_obs[2,], rev(q_obs[4,])),
        col = adjustcolor("darkgreen", 0.30),
        border = NA)

lines(x_pred, q_obs[3,], col = "darkgreen", lwd = 2)
```

:::{.callout-tip}
### EXERCISE
Extend this model to include species. Specifically, let each species have its own value of the `intercept`. This involves combining this regression example with the previous activity on discrete predictors.

When you're done, look at the resulting summary of coefficients. What do you notice that's different?
:::

:::{.callout-note collapse="true"}
### SOLUTION
```{r, include=TRUE}
#| class-output: stan
normal_regression_spp <- stan_model(file = here::here("topics/02_regression/normal_regression_spp.stan"),
                                    model_name = "normal_regression_spp")

normal_regression_spp
```

We set up a list for this model just as we did before.
Note that this time we are using TRIPLE the `pred_values`, because we want to run independent predictions for each species.


```{r}
bill_vec <- seq(min(penguins_no_NA$bill_length_center),
                max(penguins_no_NA$bill_length_center), 
                length.out = 6)

data_list_spp <- with(penguins_no_NA,
                      list(N = length(bill_length_center),
                           bill_len = bill_length_center,
                           bill_dep = bill_depth_mm,
                           spp_id = as.numeric(as.factor(species)),
                           npost = 3*6,
                           pred_values = rep(bill_vec, 3),
                           pred_spp_id = rep(1:3, each = 6)))

normal_reg_spp_post <- sampling(normal_regression_spp,
                                data = data_list_spp, refresh = 0)
```

Note that the sign of the slope is different now!

```{r}
summaryReg <- summary(normal_reg_spp_post)$summary
knitr::kable(summaryReg)
```

:::

### Plotting posterior predictions

Let's plot the posterior credible ribbons for this regression

```{r}
#| layout-ncol: 2
#| warning: false
draws_avg_spp <- rstan::extract(normal_reg_spp_post, 
                                "post_bill_dep_average")$post_bill_dep_average

draws_obs_spp <- rstan::extract(normal_reg_spp_post, 
                                "post_bill_dep_obs")$post_bill_dep_obs

spp_levels <- levels(penguins$species)
spp_colors <- c("#66C2A5", "#FC8D62", "#8DA0CB")

n_pred     <- 6  # predictions per species
x_range_spp <- range(data_list_spp$pred_values)
y_range_spp <- range(penguins_no_NA$bill_depth_mm)

# --- Average response (in one plot) --- 
plot(0, 0, 
     type = "n",
     xlim = x_range_spp, ylim = y_range_spp,
     xlab = "Bill length (centered, mm)",
     ylab = "Bill depth (mm)",
     main = "Average response")

for (s in seq_along(spp_levels)) {
  idx <- ((s - 1) * n_pred + 1):(s * n_pred)
  x_s <- data_list_spp$pred_values[idx]
  q_s <- apply(draws_avg_spp[, idx],
               2,
               quantile,
               probs = c(0.025, 0.25, 0.5, 0.75, 0.975))
  
  polygon(c(x_s, rev(x_s)), 
          c(q_s[1,], rev(q_s[5,])),
          col = adjustcolor(spp_colors[s], 0.15), 
          border = NA)
  
  polygon(c(x_s, rev(x_s)), 
          c(q_s[2,], rev(q_s[4,])),
          col = adjustcolor(spp_colors[s], 0.30),
          border = NA)
  
  lines(x_s, q_s[3,], col = spp_colors[s], lwd = 2)
  spp_pts <- penguins_no_NA[as.numeric(as.factor(penguins_no_NA$species)) == s, ]
  points(spp_pts$bill_length_center, spp_pts$bill_depth_mm,
         pch = 19, cex = 0.5, col = spp_colors[s])
}
legend("topright", legend = spp_levels, col = spp_colors, lwd = 2, bty = "n")

# --- Predicted observations (one panel per species) ---
par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (s in seq_along(spp_levels)) {
  idx <- ((s - 1) * n_pred + 1):(s * n_pred)
  x_s <- data_list_spp$pred_values[idx]
  q_s <- apply(draws_obs_spp[, idx], 2, quantile,
               probs = c(0.025, 0.25, 0.5, 0.75, 0.975))
  spp_pts <- penguins_no_NA[as.numeric(as.factor(penguins_no_NA$species)) == s, ]
  
  plot(spp_pts$bill_length_center, 
       spp_pts$bill_depth_mm,
       xlim = x_range_spp, 
       ylim = y_range_spp,
       xlab = "Bill length (centered, mm)", 
       ylab = "Bill depth (mm)",
       main = spp_levels[s], 
       pch = 19, 
       cex = 0.5, 
       col = spp_colors[s])
  
  polygon(c(x_s, rev(x_s)), 
          c(q_s[1,], rev(q_s[5,])),
          col = adjustcolor(spp_colors[s], 0.15),
          border = NA)
  
  polygon(c(x_s, rev(x_s)), 
          c(q_s[2,], 
            rev(q_s[4,])),
          col = adjustcolor(spp_colors[s], 0.30), 
          border = NA)
  
  lines(x_s, q_s[3,], 
        col = spp_colors[s], 
        lwd = 2)
}
```

## Exercise!

1. We have one model without species identity as an independent variable, and one which includes species. Look at the difference in $\sigma$ between these two models. Why did the value change?
2. **Posterior predictions** Edit the generated quantities blocks of `normal_regression.stan` and `normal_regression_spp.stan` to create replicate observations for all N observations (that is, `yrep` as we did in the model of [bill depth previously](../discrete_predictor/index.qmd)). Use this to compare the models using `bayesplot::ppc_dens_overlay()` or another function from the `bayesplot` package.


